Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2010-06-14 Text/migration.tex
Dick's changes 6/14/10
git-svn-id: http://jaga.astro.indiana.edu/paperrepo/ApJL/Migration2010@66 3b7f7548-5c66-0410-83ab-3997167635fc

{\it Migration Relative to the Disk.} The disk surface density distribution does not vary much in the asymptotic phase, so radial motion of the planet in the $t = 10$ planet run also represents radial migration relative to the gas. During the burst (e.g., Boley et al. 2006), the surface density of the disk changes dramatically. To verify that the planet is really migrating relative to the background gas disk in the $t = 0$ run, we have compared the evolution of the planet's radial position $\varpi_p(t)$ with the mass inside that radius, expressed in terms of $m_{\rm cyl}$, the fraction of the disk mass interior to the planet. Between $t = 3.52$ and $3.89$ ORP, when $\varpi_p$ increases modestly, $m_{\rm cyl}$ increases from about 0.50 to 0.53; between 3.89 and 8.47 ORP, when $\varpi_p$ decreases dramatically, $m_{\rm cyl}$ also decreases dramatically from 0.53 to 0.30. So, the planet is migrating significantly with respect to the disk mass even while the inner part of the gas disk is moving inward during the burst. For a comparison measure of the magnitude of the gas motion, $m_{\rm cyl}$ at 23 AU increases from 0.50 to 0.60 from $t =$ 3.89 to 8.47 ORP, corresponding to an gas disk inflow rate of $\sim 10^{-5} M_{\odot}/{\rm yr}$.

@@

Check against estimated type I migration without GIs. 
%Aaron: Can you do this?} 
Figure \ref{fig:a} suggests that the overall radial drift is not too different, but it is definitely not monotonic and can stall. Further work is needed to determine how long the stalls last.

@@

Discrete mode interactions with the planet seem important -- Wind up and pitch? Surfing?

Stalling at LRs?
In both cases it is interesting to note that inward migration halts near 17 AU (coincidence???) ; this may be an artifact of the length of the simulations, clearly both simulations should be carried further. This radius is interesting because it is the location of the Inner Linblad Resonance (ILR) for the strong m = 2 mode present in the baseline simulation. Inside the ILR the mode can not effectively transfer angular momentum between the disk and the planet. Further work is necessary to adequately analyze the Fourier modes and their periodicities in the simulations with planets, but this gives a strong preliminary indication that the GI activity is driving the migration. Furthermore, where GIs are inactive planet migration is stalled, even Type I migration may be ineffective due to interference from the GIs at the planet’s OLR.

Not a simple random walk for this mass and this disk. Need to know dependence on planet mass and initial location.

@@

The inclusion of a 1$M_J$ mass in the equilibrium disk initiates GIs and dramatically effects the character of the GIs. GIs can cause both inward and outward planet migration and can cause rapid migration in the burst phase.
GIs can have a strong effect on the orbital eccentricity of the included planet.

>>>>

{\it Migration Relative to the Disk.} The disk surface density distribution does not vary much in the asymptotic phase, so radial motion of the planet in the $t = 10$ planet run also represents radial migration relative to the gas. During the burst (e.g., Boley et al. 2006), the surface density of the disk changes dramatically. To verify that the planet is really migrating relative to the background gas disk in the $t = 0$ run, we have compared the evolution of the planet's radial position $\varpi_p(t)$ with the mass inside that radius, expressed in terms of $m_{\rm cyl}$, the fraction of the disk mass interior to the planet. Between $t = 3.52$ and $3.89$ ORP, when $\varpi_p$ increases modestly, $m_{\rm cyl}$ increases from about 0.50 to 0.53; between 3.89 and 8.47 ORP, when $\varpi_p$ decreases dramatically, $m_{\rm cyl}$ also decreases dramatically from 0.53 to 0.30. So, the planet is migrating significantly with respect to the disk mass even while the inner part of the gas disk is moving inward during the burst. For a comparison measure of the magnitude of the gas motion, $m_{\rm cyl}$ at 23 AU increases from 0.50 to 0.60 from $t =$ 3.89 to 8.47 ORP, corresponding to an gas disk inflow rate of $\sim 10^{-5} M_{\odot}$ yr$^{-1}$.

@@

@@

%Scott and Aaron: Can you go over this part carefully and fill in some calculation for the migration rate?
Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3 ORP, where the average radial migration is about 4$\times1o^{-3}$ AU yr$^{-1}$. Although there are no formulae that are fully applicable to our radiatively cooled disk, using equation ??? from ??? and parameters from the simulated disk, we estimate ??? AU yr$^{-1}$ as an analytic prediction. While the interaction with a GI-active disk did not seem to result in a drastically different overall migration rate, the direction and magnitude of migration in the GI-active disk is highly variable and seems to depend strongly on interactions of the planet with discrete modes in the disk. In the case of the $t = 10$ planet run, the overall radial migration rate is similar but highly erratic. The important departures from laminar disk theory are that there are times when migration stalls and that interactions with discrete low-order modes seem to be important. The migration is neither a random walk nor monotonic but has a chaotic character due to nonlinear interactions. For all these reasons, it will be important in future studies to explore the dependence of the planet's behavior on its mass and on its placement relative to the site of GI-eruptions (for $t = 0$ type runs) and relative to the CR of low-order GI modes (for $t = 10$ type runs).

%Scott and Aaron: This is REALLY speculative. We need analysis here! The following is a modified version of text from Scott's poster mixed with the idea from my dream.
In both simulations, the inward migration appears to halt near 17 AU. This may simply be a coincidence, because the overall average migration rates are similar, the starting positions are essentially the same, and the calculations have a similar duration. We plan to extend simulations like these to much longer times in the future. However, the final positions of the planets radius happens to be the location of the Inner Lindblad Resonance (ILR) for the strong m = 2 mode present in the fiducial run. In fact, this ILR is associated with a surface density enhancement in the disk. As the planet approaches this enhancement, it experiences a steeply falling surface density gradient which,
in a laminar disk, could cause the migration to stall (???). It could also be that, near the ILR, the mode no longer effectively transfers angular momentum from the planet. 
%The words that follow are from Scott. We must DO the analyses of periods, look at the surface density, look at the Q distribution (I doubt the planets are out of the GI active region yet), and think about the physics.
Further work is necessary to adequately analyze the Fourier modes and their periodicities in the simulations with planets, but this gives a strong preliminary indication that the GI activity is driving the migration. Furthermore, where GIs are inactive planet migration is stalled, even Type I migration may be ineffective due to interference from the GIs at the planet’s OLR.
%Is any of this true? An alternative but related possibility is that the planet reaches the outer edge of a surface density enhancement, so that the local density gradient is so steep the ILR's of the planet cancel the OLR's of the planet. This would probably put the planet right near the ILR of a major mode, because this is where we expect pile ups of material. (This came to me in a dream.)

@@

%Scott and Aaron: I am at a loss for a good punch line to end with. Help.
The inclusion of a 1$M_J$ mass in the equilibrium disk initiates GIs and dramatically effects the character of the GIs. GIs can cause both inward and outward planet migration and can cause rapid migration in the burst phase.
GIs can have a strong effect on the orbital eccentricity of the included planet.
